The Ranked Nearest Neighbour Interchange space

### 2019-06-27 paper.tex
some housekeeping

\documentclass[11pt, a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}

@@

\title{The Ranked Nearest Neighbour Interchange space}
\date{\today}
\author{Lena Collienne, Alex Gavryushkin}

@@

@@

\todo{citation of definitions??}

@@

Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal node $v$ gives a \emph{cluster} $C$ that is the set of taxa descending from this node. \todo{OK?}
We then say that $v$ induces the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
Such a split will be denoted by $A|B$ and we say that it is induced by edge $e$.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes

@@

One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in~\cite{Gavryushkin2017}:

@@

% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.
One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between caterpillar trees, coincide with exchanges of two neighboured elements in the sequence.
In these sequences the first two taxa are somehow equivalent since they build the unique cherry.
$\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm~\cite{knuth1997}:

@@

%Shall we put this bit into the introduction?
It is natural to try to relate the $\rnni$ graph to the $\nni$ as we already did in previous sections.
Since computing distances in tree spaces is a main issue in most commonly known tree spaces as $\nni$~\cite{jiang2000}, $\rspr$~\cite{Bordewich2005}, or $\tbr$~\cite{allen2001subtree} is $\np$-hard, it would be desirable to find a tree space that is less complex.
However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the reason that $\nni$ is $\np$-hard transfers to $\rnni$.

The proof of NP-completeness of $\nni$ in~\cite{jiang2000} is based on the fact that following theorem holds in $\nni$.

@@

\begin{proof}
	See~\cite{Li1996}.
\end{proof}

The proof presented in~\cite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$.

%This part is mainly for myself, not sure whether it is important to have it in the paper
\todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
The counterexample presented in~\cite{Li1996} works in $\nni$ because of the following:
When proving Theorem~\ref{thm:split_nni},~\cite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ where $T|_{\{1, \ldots, n\}}$ and $R|_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

Since the example given in~\cite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, the following conjecture was raised~\cite{Gavryushkin2017}.

@@

Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
Within the Cluster Theorem we are considering clusters instead of splits.
This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\cite{steel2016}, but not by sets of splits as they cannot define the position of the root.
Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.

@@

\bibliographystyle{abbrv}
\bibliography{lit}

>>>>

\documentclass{amsart}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage[foot]{amsaddr}
\usepackage[hidelinks]{hyperref}
\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
\renewbibmacro{in:}{}
\addbibresource{lit.bib}

@@

\title{The Ranked Nearest Neighbour Interchange space}
\date{\today}
\author{Lena Collienne}
\email{lena.collienne@postgrad.otago.ac.nz}
\author{Alex Gavryushkin}
\email{alex@biods.org}
\address{Department of Computer Science, University of Otago, New Zealand}

@@

Although we endeavour to make this paper self-sufficient, we refer the reader to standard textbooks in phylogenetics for terminology \autocite{Felsenstein2004-of, Semple2003-nj, Steel2016-ye}.
Unless is explicit, we follow these textbooks in our terminology.

@@

@@

Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal node $v$ gives a \emph{cluster} $C$ that is the set of taxa descending from this node. \todo{OK?}
We then say that $v$ induces the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
Such a split will be denoted by $A|B$ and we say that it is induced by edge $e$.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}:

@@

% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.
One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between caterpillar trees, coincide with exchanges of two neighboured elements in the sequence.
In these sequences the first two taxa are somehow equivalent since they build the unique cherry.
$\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{knuth1997}:

@@

%Shall we put this bit into the introduction?
It is natural to try to relate the $\rnni$ graph to the $\nni$ as we already did in previous sections.
Since computing distances in tree spaces is a main issue in most commonly known tree spaces as $\nni$ \autocite{jiang2000}, $\rspr$ \autocite{Bordewich2005}, or $\tbr$ \autocite{allen2001subtree} is $\np$-hard, it would be desirable to find a tree space that is less complex.
However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the reason that $\nni$ is $\np$-hard transfers to $\rnni$.

The proof of NP-completeness of $\nni$ in \autocite{jiang2000} is based on the fact that following theorem holds in $\nni$.

@@

\begin{proof}
	See \autocite{Li1996}.
\end{proof}

The proof presented in \autocite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$.

%This part is mainly for myself, not sure whether it is important to have it in the paper
\todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
The counterexample presented in \autocite{Li1996} works in $\nni$ because of the following:
When proving Theorem~\ref{thm:split_nni}, \autocite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ where $T|_{\{1, \ldots, n\}}$ and $R|_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

Since the example given in \autocite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, the following conjecture was raised \autocite{Gavryushkin2017}.

@@

Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in \autocite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
Within the Cluster Theorem we are considering clusters instead of splits.
This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{steel2016}, but not by sets of splits as they cannot define the position of the root.
Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.

@@

\printbibliography
